Supp mat: time series paper

### 2017-06-03 writing/manuscript_draft.tex
added to methods and intro of paper

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=0.75in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

Long-term census data is necessary to better understand ecological
processes and make management decisions. In addition, census data is
often expensive, requiring a lot of people-hours or equipment. However,
little work has actually addressed the length of time series required.
In other words, when is a time series of census data long enough to
address a question of interest? Here, I explore two approaches to
address this question: one simulation-based and an empirical approach. I
specifically determine the minimum time series length required to
estimate significant increases or decreases in population abundance.
Importantly, I examine the ability to detect a trend with a set level of
statistical power, which is often neglected in ecological time series
analyses. Using simple simulations, I demonstrate how the minimum time
series length required increases with weaker trends in abundance and
with higher variability in population size. In addition, I examine 868
populations of vertebrate species to determine the minimum time required
to detect changes in their abundance. Here I show 10-15 years of
continuous monitoring are required in order to achieve a high level of
statistical power. Similar to simulation-approaches, the minimum time
required for field census data strongly depends on trend strength,
population variability, and temporal autocorrelation. Perhaps
surprisingly, the minimum time required did not correlate well with
biological explanatory variables, like body size or generation length.
These results point to the importance of sampling populations over long
periods of time. Here, I stress the need for verifying a study has
enough statistical power to accurately determine long-term changes in
population abundance. Most studies, especially those less than 10-15
years, are probably under-powered and potentially misleading.

\subsection{Introduction}\label{introduction}

\begin{itemize}
\tightlist
\item
  intro long time series
\item
  my general question
\item
  primer on power
\item
  more general power interpretation
\item
  2+ approaches
\item
  what we do in this paper
\end{itemize}

Observational studies and population censuses have become a cornerstone
of modern ecology and conservation biology (Magurran et al. 2010).
Longterm data is necessary to understand population dynamics and to
assess extinction risk of species. Even though many time series may now
be considered ``longterm'', most are still short (i.e.~less than X
years). \emph{I could have statistic on how long is considered
``longterm'' in papers}. Time series are typically short for a variety
of reasons. They are often coupled with an experiment, which may only
last a couple of years. In addition, short funding cycles make it
difficult to examine populations over longer periods of time. There are
of course notable exceptions including the Isle Royale study, Iowa corn
study, cod, lynx-hair, and many others.

Naturally, one might ask, how long of a time series is actually
necessary? This question has important scientific and management
implications. Scientifically, we need to know how long a time series is
required to address a specific question. Too short of a time series may
lead to wrong conclusions given large natural year-to-year variability.
From a management perspective, it is important to understand when a
trend in abundance over time is actually meaningful or not. For instance
the International Union for the Conservation of Nature (IUCN)
suggest\ldots{}In addition, because sampling is typically expensive, we
do not want to over sample (Gerber, DeMaster, and Kareiva 1999).

In determining the number of samples required for a particular
experiment four quantities are intricately linked: statistical power,
effect size, alpha, sample size (Legg and Nagy 2006). The exact
relationship between these quantities depends on the specific
statistical test. Formally, statistical power is one minus the
probability of a type II error, or false negative. In other words, it is
a measure of how well you could detect a trend given it is actually
significant. Prior to an experiment, one could set appropriate levels of
power, \(\alpha\), and the effect size to estimate the sample size
required for the experiment. This approach, however, is not as simple
for a time series as the data points are clearly non-independent.

For time series data, two general approaches can be used. First, as can
be done with experimental data, we can relax the definition of
statistical power (cite Bolker). Instead of using classic statistical
tests (e.g.~t-test), and corresponding power calculations, simulation
approaches can be used (Gerrodette 1987). Simple population models can
be simulated with parameter values corresponding to a population of
interest. Out of a large number of simulations, power can be calculated
as the fraction of simulations that meet some criteria. The specific
criteria depend on the question at hand. Then one can run simulations
for time series of different lengths to determine the minimum time
series length required to be confident that you will be able to answer
your question. In addition to using simulated data, empirical time
series can also be used. We usually do not have multiple replicates of
similar populations, but we can subsample an empirical time series.
Subsamples of different lengths can then be evaluated to see which
fraction of subsamples meet some criteria, again a measure of
statistical power\ldots{}

Past work has investigated these questions with either theoretical
approaches not using data or by examining a single population. As an
example of the former, (Rhodes and Jonzen 2011) examined.. - need to
expand this paragraph in line with what I wrote in the
discussion\ldots{}

@@

One approach to determining the minimum time series needed is through
repetitive simulations of a population model (Gerrodette 1987). This is
the same approach one might use in sample size calculations for any
experimental design too complicated for simple power analyses (Johnson
et al. 2015). We only briefly discuss this approach as it has been
desribed elsewhere (citations). Essentially, we take a stochastic
population model and simulate it for a number of years repetitively. We
can then determine the minimum time series required to achieve designed
levels of significance and power. This approach requires us to determine
values for our model parameters. As an example, we can take the
following population model for population size \(N\) at time \(t\): \[
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\phi \mu, \sigma)
\] where \(\epsilon\) is a normally-distributed random noise term with
mean \(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\)
is also the trend strength of increase or decrease. Lastly, \(\phi\)
represents an autocorrelation term where \(\phi=0\) indicates no
autocorrelation, \(\phi>0\) is positively autocorrelated. This model
represents linear growth with autocorrelation and noise. However, it is
certainly possible to use any other population model.

By simulating this model repetitively for different lengths of time, we
can determine the minimum time series required. Here statistical power
is simply the number of trials that are themselves significant, as the
real trend is different from zero. Then the minimum time required is the
minimum time required to obtain statistical power above some
pre-determined threshold. Historically, 0.8\ldots{}.s

In figure X, a number of simulated time series are shown for a set
number of time periods (t=40). It is clear that statistical power
increases quickly with increases in \(t\) (Fig. Xb). Once power is
greater than XX (the dotted line), that is the minimum time required
(\(t_{min}\)

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{theoretical approach\label{fig:theoretical_approach}}
\end{figure}

@@

We use a database of 2444 population time series compiled in (Keith et
al. 2015). The data is originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010). We filtered out
short time series (less than 35) and were left with 1019 time. The
database includes information on 499 vertebrate species with a focus on
mammals, birds, and fish. The database also includes information on
generation length and census specifications. For each time series, we
also calculated variables of interest like variance in population size,
longterm trend in abundance (slope coefficient from simple linear
regression), and temporal autocorrelation. Lastly, we also calculated
the minimum number of years required to sample according to IUCN
standards under rule \ldots{} (citation needed).

@@

Our objective is to determine the minimum time series length
(\(T_{min}\)) required to answer a specific question of interest. As we
describe below, the minimum time series length calculation is particular
to each question. Intuitively, we might expect a longer minimum time
series length required given more complicated questions.

We assume that each time series in our database are long enough to
include all necessary information (e.g.~variability) about the
population. In other words, they are a Representative sample. We first
take all possible contiguous subsamples of each time series. For
example, a time series of 40 years would have 39 possible contiguous
subsamples of length 2, 38 possible contiguous subsamples of length 3,
and so forth. In this paper, we will try to ask a simple, specific
question. We seek to determine the \(T_{min}\) required to be confident
in an estimate of the slope coefficient from linear regression.
Therefore, we run a linear regression for each subsampled time series.
Then, we determine the fraction of subsamples of a particular length
that have slopes from linear regression which are statistically
different from zero given the longterm, or ``true``, time series also
has a significant slope. This is a measure of statistical power. Lastly,
we determine which subsample length is required to guarantee we have
above a certain threshold of statistical power, which here we will
assume is 0.8 by convention (citation).

We also examine how minimum time series calculations changed depending
on the question of interest. In the supplementary material, we show how
to make similar calculations to determine the minimum time series
required to estimate overall population growth rate, the variability in
population size, and future extinction risk.

In the appendix, we explore other potential questions of
interest\ldots{}

@@

\begin{itemize}
\tightlist
\item
  summary of findings
\item
  theoretical approach
\item
  example plot
\item
  empirical approach
\item
  plots: distribution of min time, what min time correlations with
\item
  walk through one empirical example (maybe in appendix)
\end{itemize}

@@

We constructed a general population model where the trend over time
could be a model parameter. We chose a simple model, but any other
population model could be used. The specific model choice should be
tailored to the population of interest, based on species ecology. We
then ran simulations of different lengths to determine the minimum time
series length required to achieve a certain level of power (which we set
to 0.8). We were also able to alter the parameter values (trend strength
and population variability) of the population model to see how this
altered \(T_{min}\).

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Distribution of the minimum time required in order to detect a
significant trend (at the 0.05 level) in abundance given power of
0.8.\label{fig:fig:min_time_dist}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{duh\label{fig:correlates}}
\end{figure}

@@

Across all the populations we examined, we found an average minimum time
series length required (\(T_{min}\)) of 16.6463134
(\(\sigma\)=8.6664727), with a wide distribution (Fig.
\ref{fig:min_time_dist}). As expected based on theoretical results (cite
papers), the minimum time series length required is strongly correlated
with trend strength, population variance, and autocorrelation (Fig
\ref{fig:correlates}). Using a generalized linear model, with a Poisson
error structure, all three of these explanatory variables were
significant and had large effect sizes (cite model here). Combined trend
strength, population variance, and autocorrelation account for about
70.7793122 of the variance in minimum time series length required (as
proposed by CITE Zuur).

We also examined a subset of populations where more explanatory
variables were known (e.g.~body mass, longevity, location). None (need
to do regression) of these explanatory variables were predictive of the
minimum time series length required. Some differences if \(t_{min}\) did
occur between biological classes, but this can be attributed to the
amount populations are know to fluctuate.

@@

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

Long-term census data is necessary to better understand ecological
processes and make management decisions. In addition, census data is
often expensive, requiring a lot of people-hours or equipment. However,
little work has actually addressed the length of time series required.
In other words, when is a time series of census data long enough to
address a question of interest? Here, I explore two approaches to
address this question: one simulation-based and an empirical approach. I
specifically determine the minimum time series length required to
estimate significant increases or decreases in population abundance.
Importantly, I examine the ability to detect a trend with a set level of
statistical power, which is often neglected in ecological time series
analyses. Using simple simulations, I demonstrate how the minimum time
series length required increases with weaker trends in abundance and
with higher variability in population size. In addition, I examine 868
populations of vertebrate species to determine the minimum time required
to detect changes in their abundance. Here I show 10-15 years of
continuous monitoring are required in order to achieve a high level of
statistical power. Similar to simulation-approaches, the minimum time
required for field census data strongly depends on trend strength,
population variability, and temporal autocorrelation. Perhaps
surprisingly, the minimum time required did not correlate well with
biological explanatory variables, like body size or generation length.
These results point to the importance of sampling populations over long
periods of time. Here, I stress the need for verifying a study has
enough statistical power to accurately determine long-term changes in
population abundance. Most studies, especially those less than 10-15
years, are probably underpowered and potentially misleading.

\subsection{Introduction - need to clean up and reduce
text}\label{introduction---need-to-clean-up-and-reduce-text}

Observational studies and population censuses have become a cornerstone
of modern ecology and conservation biology (Magurran et al. 2010).
Longterm data is necessary to understand population dynamics and to
assess extinction risk of species. Even though many time series may now
be considered ``longterm'', most are still short (i.e.~less than X
years). \emph{I could have statistic on how long is considered
``longterm'' in papers}. Time series are typically short for a variety
of reasons. They are often coupled with an experiment, which may only
last a couple of years. In addition, short funding cycles make it
difficult to examine populations over longer periods of time. However,
there are of course notable exceptions including the Isle Royale study,
Iowa corn study, cod, lynx-hair, and many others.

Naturally, one might ask, how long of a time series is actually
necessary? This question has important scientific and management
implications. Scientifically, we need to know how long a time series is
required to address a specific question. Too short of a time series may
lead to wrong conclusions given large natural year-to-year variability
(citations). From a management perspective, it is important to
understand when a trend in abundance over time is actually meaningful or
not. For example, the IUCN Red List Categories and Criteria suggest
under Criterian A2, a species qualifies as vulnerable if it has
experienced a 30\% decline over 10 years, or 3 generations (IUCN 2012).
In addition, because sampling is typically expensive, we do not want to
sample for longer than is neccessary (Gerber, DeMaster, and Kareiva
1999).

In determining the number of samples required for a particular
experiment four quantities are intricately linked: statistical power,
effect size, alpha, sample size (Legg and Nagy 2006). The exact
relationship between these quantities depends on the specific
statistical test. Formally, statistical power (\(1-\beta\)) is one minus
the probability of a type II error (\(\beta\)), or false negative. In
other words, it is a measure of how well you could detect a trend given
it is actually significant. Prior to an experiment, one could set
appropriate levels of power, significance threshold (\(\alpha\)), and
the effect size to estimate the sample size required for the experiment.
This approach, however, is not straight-forward for a time series, as
data points are clearly non-independent.

For time series data, two general approaches can be used. First, as can
be done with experimental data, we can relax the definition of
statistical power (cite Bolker). Instead of using classic statistical
tests (e.g.~t-test), and corresponding formulas for determinig sample
size, simulation approaches can be used (Gerrodette 1987). Simple
population models can be simulated with parameter values corresponding
to a population of interest. Out of a large number of simulations, power
can be calculated as the fraction of simulations that meet some
criteria. The specific criteria depend on the question at hand. Then
time series simulations of different lengths are run to determine the
minimum time series length required to be confident that you will be
able to answer your question (cite Bolker). In addition to using
simulated data, empirical time series can also be used. We usually do
not have multiple replicates of similar populations, but we can
subsample an empirical time series (cite). Subsamples of different
lengths can then be evaluated to see which fraction of subsamples meet
some criteria, again a measure of statistical power. Similar to the
simulation approach, we can use these measures of power to determine the
minimum time series required for a particular question of interest.

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time. However, most
work has either focused on theoretical approaches or by examined a
single population. As an example of the former, Rhodes and Jonzen (2011)
examined the optimal allocation of effort between spatial and temporal
replicates. Using simple populations models, they found that the
allocation of effort depends on environmental variation, spatial and
temporal autocorrelation, and observer error. Rueda-Cediel et al. (2015)
also used a modeling approach, but parameterized it for a threatened
snail, \emph{Tasmaphena lamproides}. They found that for this
short-lived organims, that 15 years was adequate to assess longterm
trends in abundance. However, these studies, and other past work
(citation), have focused either on theoretical aspects of monitoring
design or focused on single species.

@@

One approach to determining the minimum time series length needed is
through repetitive simulations of a population model (Gerrodette 1987).
This is the same approach one might use in sample size calculations for
any experimental design too complicated for simple power analyses
(Johnson et al. 2015) (and cite Bolker). We only briefly discuss this
approach as it has been desribed elsewhere (citations). Essentially, we
take a stochastic population model and simulate it for a number of years
repetitively. We can then determine the minimum time series required to
achieve designed levels of significance and power for a particular
question. This approach requires us to determine values for our model
parameters (e.g.~growth rate). As an example, we can take the following
population model for population size \(N\) at time \(t\): \[
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\phi \mu, \sigma)
\] where \(\epsilon\) is a normally-distributed random noise term with
mean \(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\)
is also the trend strength of increase or decrease.

By simulating this model repetitively for different lengths of time, we
can determine the minimum time series required. Here statistical power
is simply the fraction of simulations with significant trends in
abundance from simple linear regression. Then, the minimum time required
is the minimum time required to obtain statistical power above some
pre-determined threshold (historically at 0.8).

In figure \ref{fig:theoretical_approach}a, a number of simulated time
series are shown for a set number of time periods (t=40). It is clear
that statistical power increases quickly with increases in \(t\) (Fig.
\ref{fig:theoretical_approach}b). Once power is greater than XX (the
dotted line), that is the minimum time required (\(t_{min}\)

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{(a) example of simulated time series for 40 time periods and
(b) statistical power versus the simulated time series length. The
horizontal, dashed line is our desired statistical power of 0.8. The
vertical, dashed line is the minimum time required to achieve our
desired statistical power.\label{fig:theoretical_approach}}
\end{figure}

@@

We use a database of 2444 population time series compiled in (Keith et
al. 2015). The data is originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010). We filtered out
short time series (less than 35) and were left with 1019 time. The
database includes information on 499 vertebrate species with a focus on
mammals, birds, and fish. The database also includes information on
generation length and census specifications. For each time series, we
also calculated variables of interest like variance in population size,
longterm trend in abundance (slope coefficient from simple linear
regression), and temporal autocorrelation. Lastly, we also calculated
the minimum number of years required (10 years or 3 generations) under
IUCN Red List Criterian A2 to classify a species as vulnerable (IUCN
2012).

@@

Our objective is to determine the minimum time series length
(\(T_{min}\)) required to answer a specific question of interest. As we
describe below, the minimum time series length calculation is particular
to each question. Intuitively, we might expect a longer minimum time
series length (\(T_{min}\)) required given more complicated questions
(citations).

We assume that each time series in our database is long enough to
include all necessary information (e.g.~variability) about the
population. In other words, they are a representative sample. We first
take all possible contiguous subsamples of each time series. For
example, a time series of 40 years would have 39 possible contiguous
subsamples of length 2, 38 possible contiguous subsamples of length 3,
and continueing until 1 possible contiguous subsample of length 40. In
this paper, we will try to ask a simple, specific question. We seek to
determine the \(T_{min}\) required to be confident in an estimate of the
slope coefficient from linear regression. Therefore, we run a linear
regression for each subsampled time series. Then, we determine the
fraction of subsamples of a particular length that have estimated slope
coefficients which are statistically different from zero given the
longterm, or ``true``, time series also has a significant slope. This is
a measure of statistical power (often denoted as \(1- \beta\), where
\(\beta\) is the probability of a type II error). Lastly, we determine
which subsample length is required to guarantee we have above a certain
threshold of statistical power, which we will assume is 0.8 by
convention (citation). A power of 0.8 would imply that given a trend
exists, there is a 0.8 probability of detecting the trend.
Complementary, there would be a 0.2 probability of a false negative.

We also examine how minimum time series calculations changed depending
on the question of interest. In the supplementary material, we show how
to make similar calculations to determine the minimum time series
required to estimate overall population growth rate, the variability in
population size, and future extinction risk. Again, we took subsamples
of time series data to determine \(T_{min}\) for each of these
questions.

@@

@@

We constructed a general population model where the trend over time
could be a model parameter. We chose a simple model, but any other
population model could be used. The specific model choice should be
tailored to the population of interest, based on its ecology. We then
ran simulations of different lengths to determine the minimum time
series length required to achieve a certain level of power (which we set
to 0.8). We were also able to alter the parameter values (trend strength
and population variability) of the population model to see how this
altered \(T_{min}\).

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Distribution of the minimum time required in order to detect a
significant trend (at the 0.05 level) in abundance given power of
0.8.\label{fig:min_time_dist}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Minimum time required to estimate change in abundance
correlated with (a) trend strength (absolute value of slope coefficient
estimated from linear regression), (b) population variance (interannual
variability in population size), and (c) temporal
autocorrelation.\label{fig:correlates}}
\end{figure}

@@

Across all the populations we examined, we found an average minimum time
series length required (\(T_{min}\)) of 16.6 (\(\sigma\)=8.7), with a
wide distribution (Fig. \ref{fig:min_time_dist}). In addition, as
expected based on theoretical results (cite papers), the minimum time
series length required is strongly correlated with trend strength,
population variance, and autocorrelation (Fig \ref{fig:correlates}).
Using a generalized linear model, with a Poisson error structure, all
three of these explanatory variables were significant and had large
effect sizes (cite model from supp mat here). Combined trend strength,
population variance, and autocorrelation account for about 70.8 percent
of the variance in minimum time series length required (as proposed by
CITE Zuur).

We also examined a subset of populations where more explanatory
variables were known (e.g.~body mass, longevity, location). None (need
to do regression) of these explanatory variables were predictive of the
minimum time series length required (supp mat). Some differences if
\(t_{min}\) did occur between biological class (supp mat), but this can
be attributed to the amount populations are known to fluctuate (i.e.~the
variability in population size).

@@

\hypertarget{ref-IUCN2012}{}
IUCN. 2012. ``IUCN Red List Categories and Criteria: Version 3.1.''
doi:\href{https://doi.org/10.9782-8317-0633-5}{10.9782-8317-0633-5}.
